Comparing R packages for calculation of persistent homology

### 2020-03-19 tdabench-paper/tdabench.Rmd
alpha capitalization

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. The Dionysus (TDA), GUDHI (TDA), and Ripser C++ libraries (TDAstats) provide Rips complex functionality. GUDHI  (TDA package) also provides Alpha complex functionality. These software represent powerful analysis tools but are limited computationally. These limitations and guidelines for which persistent homology libraries to use have not been specified in R. We benchmarked the all libraries on a High-Powered Cluster for runtime. We also compared Alpha and Rips for memory burden. Alpha complexes run significantly faster than all Rips complex libraries for all data sets for data dimension less than or equal to three. Alpha complexes were not computable for data dimension greater than 3. TDAstats had the fastest runtimes for all Rips complexes. We recommend performing persistent homology with GUDHI Alpha complexes for data dimension less than or equal to three. For higher dimension, we recommend TDAstats Rips Complex. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

@@

In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize the topological features. 
Simplicial complexes are connected sets of points, lines, and n-dimensional shapes.
There are several methods to generate these complexes on point clouds. 
In this paper, we are using the Rips and Alpha complex for our persistent homology calculations. 
In both methods, a series of data structures, called simplicial complexes, are built on top of a point cloud.
These complexes can be viewed as approximating real topologic relationships in the data.
The exact method of constructing these complexes is described in the Mathematics section. 
However, the key takeaway is that the simplicial complex attempts to reveal the underlying topology of the point cloud. 
These features can be described by dimension and "persistence." 
Long lasting (persistent) features that exist across the series of simplicial complexes intuitively represent "real" geometrical patterns as opposed to noise.

@@

There are several C++ libraries that calculate Alpha and Rips complexes.
Dionysus provides Alpha and Rips complex functionality \cite{Dionysus}.
GUDHI is another C++ library that also provides both functionalities as well as other TDA methodologies \cite{gudhi}. These libraries have been implemented in R with the TDA package \cite{TDA} providing an easier interface to utilize persistent homology calculations.

@@

Topological data analysis has had a broad range of usage from viral evolution to physical chemistry\cite{TDA-Viral,TDA-PChem}. 
Given that persistent homology can apply to any n-dimensional data set, it has wide ranging applicability. 
Its main limiting factor is computational for calculating high dimensional features in large multi-dimensional data sets. 
We believe lack of analysis of these higher dimensional features may under utilize persistent homology's analysis potential.
In this project, we explore different methodologies to find efficient options that could solve this issue. 
We specifically aim to characterize the computational performance (run time) and limits of persistent homology calculations between Rips complexes in the R packages TDAstats and TDA.
We also explore the differences in computational performance between Rips and Alpha complexes in R to assess which datasets are most appropriate for each algorithm.

@@

There are several different methods to construct a simplicial complex on a given point cloud S, but this paper focuses on the Rips and Alpha complexes. 
The Rips complex is perhaps the most common method for constructing a simplicial complex under persistent homology \cite{Rips-Complex}.
In a point cloud of k points, a distance parameter, $\delta$, can be used to draw a circle of diameter $\delta$ around every single point in S.
If the distance parameter is non-zero, then some of the resulting circles may intersect.
If they intersect, a line is drawn to connect the points (which represent the centers of the growing circles).
This group of points and lines form a simplicial complex.
For each distance parameter, $\delta$, there will be some simplicial complex associated with it.
As $\delta$ increases, different topological features may appear, persist, then disappear once $\delta$ has reached a threshold.
Eventually, once $\delta$ reaches the maximum euclidean distance between any pair of points in the point cloud, a convex hull will form around all k points creating a (k-1)-dimensional simplex.
At this stage, all features should have disappeared.
A three column matrix can be created recording the Betti number of a certain feature, the distance, $\delta$, at which that feature appeared, and the $\delta$ at which that feature disappeared.
This matrix characterizes the persistent homology of that point cloud.

@@

In both methods, the boundary matrix records all simplicial complexes for a given parameter value. 
From this matrix, the persistent homology matrix is calculated. 
The size of the boundary matrix can describe the memory complexity on the RAM for persistent homology calculations. 
We compare this memory complexity between Alpha and Rips complexes for this project.

>>>>

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. The Dionysus (TDA), GUDHI (TDA), and Ripser C++ libraries (TDAstats) provide Rips complex functionality. GUDHI  (TDA package) also provides alpha complex functionality. These software represent powerful analysis tools but are limited computationally. These limitations and guidelines for which persistent homology libraries to use have not been specified in R. We benchmarked the all libraries on a High-Powered Cluster for runtime. We also compared alpha and Rips for memory burden. Alpha complexes run significantly faster than all Rips complex libraries for all data sets for data dimension less than or equal to three. alpha complexes were not computable for data dimension greater than 3. TDAstats had the fastest runtimes for all Rips complexes. We recommend performing persistent homology with GUDHI alpha complexes for data dimension less than or equal to three. For higher dimension, we recommend TDAstats Rips Complex. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

@@

In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize the topological features. 
Simplicial complexes are connected sets of points, lines, and n-dimensional shapes.
There are several methods to generate these complexes on point clouds. 
In this paper, we are using the Rips and alpha complex for our persistent homology calculations. 
In both methods, a series of data structures, called simplicial complexes, are built on top of a point cloud.
These complexes can be viewed as approximating real topologic relationships in the data.
The exact method of constructing these complexes is described in the Mathematics section. 
However, the key takeaway is that the simplicial complex attempts to reveal the underlying topology of the point cloud. 
These features can be described by dimension and "persistence." 
Long lasting (persistent) features that exist across the series of simplicial complexes intuitively represent "real" geometrical patterns as opposed to noise.

@@

There are several C++ libraries that calculate alpha and Rips complexes.
Dionysus provides alpha and Rips complex functionality \cite{Dionysus}.
GUDHI is another C++ library that also provides both functionalities as well as other TDA methodologies \cite{gudhi}. These libraries have been implemented in R with the TDA package \cite{TDA} providing an easier interface to utilize persistent homology calculations.

@@

Topological data analysis has had a broad range of usage from viral evolution to physical chemistry\cite{TDA-Viral,TDA-PChem}. 
Given that persistent homology can apply to any n-dimensional data set, it has wide ranging applicability. 
Its main limiting factor is computational for calculating high dimensional features in large multi-dimensional data sets. 
We believe lack of analysis of these higher dimensional features may under utilize persistent homology's analysis potential.
In this project, we explore different methodologies to find efficient options that could solve this issue. 
We specifically aim to characterize the computational performance (run time) and limits of persistent homology calculations between Rips complexes in the R packages TDAstats and TDA.
We also explore the differences in computational performance between Rips and alpha complexes in R to assess which datasets are most appropriate for each algorithm.

@@

There are several different methods to construct a simplicial complex on a given point cloud S, but this paper focuses on the Rips and alpha complexes. 
The Rips complex is perhaps the most common method for constructing a simplicial complex under persistent homology \cite{Rips-Complex}.
In a point cloud of k points, a distance parameter, $\delta$, can be used to draw a circle of diameter $\delta$ around every single point in S.
If the distance parameter is non-zero, then some of the resulting circles may intersect.
If they intersect, a line is drawn to connect the points (which represent the centers of the growing circles).
This group of points and lines form a simplicial complex.
For each distance parameter, $\delta$, there will be some simplicial complex associated with it.
As $\delta$ increases, different topological features may appear, persist, then disappear once $\delta$ has reached a threshold.
Eventually, once $\delta$ reaches the maximum euclidean distance between any pair of points in the point cloud, a convex hull will form around all k points creating a (k-1)-dimensional simplex.
At this stage, all features should have disappeared.
A three column matrix can be created recording the Betti number of a certain feature, the distance, $\delta$, at which that feature appeared, and the $\delta$ at which that feature disappeared.
This matrix characterizes the persistent homology of that point cloud.

@@

In both methods, the boundary matrix records all simplicial complexes for a given parameter value. 
From this matrix, the persistent homology matrix is calculated. 
The size of the boundary matrix can describe the memory complexity on the RAM for persistent homology calculations. 
We compare this memory complexity between alpha and Rips complexes for this project.
